Generalizing the Model View Controller Paradigm

### 2006-08-09 inst/paper/MVC.tex
Added new versions of papers and added paper 2 version 2 (paper2vs.tex) that is my current version to give to the committee.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@19090 bc3139a8-67e5-0310-9ffc-ced21a209358

%% also this paper: Interactive data visualization using focusing and linking
%%Full text 	pdf formatPdf (541 KB)
%%Source 	IEEE Visualization archive
%%Proceedings of the 2nd conference on Visualization '91 table of contents
%%San Diego, California
%%SESSION: Multivariate visualization table of contents
%%Pages: 156 - 163  
%%Year of Publication: 1991
%%ISBN:0-8186-2245-8 (PAPER) 
There are many ways to visualize multivariate data when more than
two or three dimensions are needed to view the data.  Many authors
\citep{intGrUnwin, iPlots, MONDRIAN, GGobi}
have suggested that linked, interactive views of the data
provide a valuable and important set of tools for comprehending such
data.  The model view controller (MVC) paradigm is now a standard
mechanism for implementing such systems.  The paradigm has proven very
effective and allows for clear separation of different components.
In this paper we consider an extension of the MVC paradigm to allow
for, what we believe to be, an easy expansion to linked data sets.

@@

%%FIXME: need to reference iPlots from Simon, and some of these refs need
%%to be put up at the front where I have a citations in parens
Linked views, which have a long history in statistics, are now
considered a standard feature of interactive data visualization
software \cite{GGobi}.  Linked views have been implemented in several
pieces of statistical software, such as XLisp-Stat \cite{Lisp} and
GGobi \cite{GGobiMan}.  Implementing linked views is based on the
model-view-controller (MVC) design \cite{DesignPatterns}, which is a
widely used and well understood paradigm.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=5in, width=6in]{MultipleMVC.jpg}
    \caption{ Example of Two \Robject{MVC} Objects.  These two \Robject{MVC}
      objects are linked by the StateCity table and the color of the points
    indicates the linking between the data.  The points referring to the state
    of Pennsylvania are red, the points referring to the state of California
    are blue, and the points referring to Minnesota are green. }
    \label{Fig:MultMVC}
  \end{center}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=4in, width=5in]{newViewClass.jpg}
    \caption{ Inheritance for View Objects. }
    \label{Fig:View}
  \end{center}
\end{figure}

@@

All views need an update method so they can alter themselves when the
model sends them a message that the data have been altered. The
actions taken depend on the view, for example for a spreadsheet view
the method may select or unselect a row on the spreadsheeet while for
a scatter plot view the method may change the appearance of a point on
the plot.  The inheritance structure of the view classes is
extensible.  For instance, if a user wanted to add a new plot view,
say a density plot, they can create a new view class that inherits
from \Rclass{plotView}, perhaps called \Rclass{dPlotView}.

@@

Deciding when new data should be added to the current model versus generating
a new submodel in a \Robject{MVC} instance is done by comparing the
relationship between the new data and the data in the current model.  If the
new data has a one-to-one relationship with the elements in the current model,
then we believe that in most cases the new data should be added to the 
current model.  For example, if there is
a test statistic for each row in a data frame, then this could be added to the
model.  However, if the new data has a one-to-many or a many-to-many
relationship with the elements in the current model, then serious consideration should be given to the creation of a submodel.  If there is already an appropriate MVC class (e.g. if the submodel is a graph) then it can be
generated by creating a new instance of the appropriate MVC class.  If there
is no appropriate implementation, then the user will need to create one.  
As an example, consider the
relationship between genes and GO terms.  If the original model contains gene
expression levels and a user wants to explore the GO terms for these genes,
then we encourage the creation of the GO graph representing the GO terms,  and
creating a new graph \Robject{MVC} object.
%%FIXME: maybe explicitly name the class in the above line

@@

The functionality that is provided in the \Rpackage{iSNetwork} package is
discussed in detail in the Vignette for this R package.  Users can load data,
create views of the data, interact with these views and change the response to
the interaction (these interactions include mouse button click events and
mouse movement events), and users can create child \Robject{MVC} objects
through the GUI or through command line functions.  With the functionality in
the \Rpackage{iSNetwork} package, the user can create views of linked models
and can explore visually the relationships between variables in different data
sets.  This functionality was implemented by extending the MVC paradigm in a
very natural way to encompass a much richer class of models and user
interactions.

>>>>

%% also this paper: Interactive data visualization using focusing and linking
%%Full text     pdf formatPdf (541 KB)
%%Source        IEEE Visualization archive
%%Proceedings of the 2nd conference on Visualization '91 table of contents
%%San Diego, California
%%SESSION: Multivariate visualization table of contents
%%Pages: 156 - 163  
%%Year of Publication: 1991
%%ISBN:0-8186-2245-8 (PAPER)

%% EAW: put new citations below
There are many ways to visualize multivariate data when more than
two or three dimensions are needed to view the data.  Many authors
\citep{intGrUnwin, DynGraphics, IEEEVisual, GGobi, DataDesk}
have suggested that linked, interactive views of the data
provide a valuable and important set of tools for comprehending such
data.  The model view controller (MVC) paradigm is now a standard
mechanism for implementing such systems.  The paradigm has proven very
effective and allows for clear separation of different components.
In this paper we consider an extension of the MVC paradigm to allow
for, what we believe to be, an easy expansion to linked data sets.

@@

%%FIXME: need to reference iPlots from Simon, and some of these refs need
%%to be put up at the front where I have a citations in parens
%%EAW: added a reference to iPlots below
Linked views, which have a long history in statistics, are now
considered a standard feature of interactive data visualization
software \cite{GGobi}.  Linked views have been implemented in several
pieces of statistical software, such as XLisp-Stat \cite{Lisp}, iPlots
\cite{iPlots}, and
GGobi \cite{GGobiMan}.  Implementing linked views is based on the
model-view-controller (MVC) design \cite{DesignPatterns}, which is a
widely used and well understood paradigm.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=5in, width=6in]{MultipleMVC.jpg}
    \caption{ Example of two \Robject{MVC} objects.  These two \Robject{MVC}
      objects are linked by the StateCity table and the color of the points
    indicates the linking between the data.  The points referring to the state
    of Pennsylvania are red, the points referring to the state of California
    are blue, and the points referring to Minnesota are green. }
    \label{Fig:MultMVC}
  \end{center}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=4in, width=5in]{newViewClass.jpg}
    \caption{ Inheritance for view objects. }
    \label{Fig:View}
  \end{center}
\end{figure}

@@

All views need an update method so they can alter themselves when the
model sends them a message that the data have been altered. The
actions taken depend on the view, for example for a spreadsheet view
the method may select or unselect a row on the spreadsheet while for
a scatter plot view the method may change the appearance of a point on
the plot.  The inheritance structure of the view classes is
extensible.  For instance, if a user wanted to add a new plot view,
say a density plot, they can create a new view class that inherits
from \Rclass{plotView}, perhaps called \Rclass{dPlotView}.

@@

Deciding when new data should be added to the current model versus generating
a new submodel in a \Robject{MVC} instance is done by comparing the
relationship between the new data and the data in the current model.  If the
new data has a one-to-one relationship with the elements in the current model,
then we believe that in most cases the new data should be added to the 
current model.  For example, if there is
a test statistic for each row in a data frame, then this could be added to the
model.  However, if the new data has a one-to-many or a many-to-many
relationship with the elements in the current model, then serious
consideration should be given to the creation of a submodel.  If there is
already an appropriate MVC class (e.g. if the submodel is a graph) then it can
be generated by creating a new instance of the appropriate MVC class.  If there
is no appropriate implementation, then the user will need to create one.  
As an example, consider the
relationship between genes and GO terms.  If the original model contains gene
expression levels and a user wants to explore the GO terms for these genes,
then we encourage the creation of the GO graph representing the GO terms, and
creating a new graph \Robject{MVC} object.
%%FIXME: maybe explicitly name the class in the above line
%% EAW: Not sure what you mean about naming the class: creating the GO graph
%% creates an object of class graphModel that is placed in the model slot of a
%% linkedModelMVC object.  As for naming the model (MVC), the user must decide
%% the name.

@@

The functionality that is provided in the \Rpackage{iSNetwork} package, which
is an implementation of the research discussed in this paper, is
discussed in detail in the vignette for this R package.  Users can load data,
create views of the data, interact with these views and change the response to
the interaction (these interactions include mouse button click events and
mouse movement events), and users can create child \Robject{MVC} objects
through the GUI or through command line functions.  With the functionality in
the \Rpackage{iSNetwork} package, the user can create views of linked models
and can explore visually the relationships between variables in different data
sets.  This functionality was implemented by extending the MVC paradigm in a
very natural way to encompass a much richer class of models and user
interactions.

\section*{Acknowledgements}
We are indebted to Dr. Duncan Temple Lang for the creation of the
\Rpackage{RGtk} and \Rpackage{gtkDevice} packages as well as helpful advice on
their use.

### 2006-08-09 inst/paper/paper2.tex
Added new versions of papers and added paper 2 version 2 (paper2vs.tex) that is my current version to give to the committee.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@19090 bc3139a8-67e5-0310-9ffc-ced21a209358

\usepackage{epsfig}
\usepackage{comment}
\usepackage{graphics}
\usepackage{graphicx}

\newcommand{\Rfunction}[1]{{\textsf{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}
\newcommand{\Rslot}[1]{\textsl{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}

@@

\begin{titlepage}
\begin{center}
\Large
\vspace*{1in}
\textbf{Creating Linked, Interactive Views to \linebreak Explore Linked Data} \\
\vspace{1in}
\textbf{Dana-Farber Cancer Institute}\\ 
\textbf{May 6, 2006} \\
\vspace{1in} 
\textbf{Elizabeth Whalen} \\
\end{center}
\end{titlepage}
\normalsize
\newpage

%\renewcommand{\baselinestretch}{1.5}
\section{Introduction}\label{Sec:Intro}

@@

%\clearpage

@@

As with the model classes, the view classes need an update
method and thus, all views have an \Rfunction{updateView} method that is
defined in the \Rpackage{iSNetwork} package.  This method must take the
information that a model has just been updated and appropriately update the
view to reflect that change.  Each method depends on how the view
represents the data so for a \Robject{spreadView} object the
\Rfunction{updateView} method may select or unselect a row on the spreadsheeet
while for a \Robject{sPlotView} object the \Rfunction{updateView} method may
change the appearance of a point on the plot.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{newMessageClass2.jpg}}
    \caption{ Inheritance for Message Classes that are Passed Between
      \Robject{MVCs}. }
    \label{Fig:BetwMess}
  \end{center}
\end{figure}

@@

>>>>

\usepackage{epsfig}
\usepackage{comment}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{times}
\usepackage[authoryear,round]{natbib}
\usepackage{subfigure}

\newcommand{\Rfunction}[1]{{\textsf{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}
\newcommand{\Rslot}[1]{\textsl{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}
\renewcommand{\baselinestretch}{1.5}

@@

\author{Elizabeth Whalen\\Harvard School of Public Health}
\title{Generalizing the Model View Controller Paradigm}

\maketitle

\begin{abstract} 
  In this paper we consider a generalization of the model
  view controller (MVC) paradigm to allow for more flexible handling
  of submodels.  Instead of using a single MVC we develop a tree
  structure of related MVCs, where each MVC has a single parent, but a
  parent can spawn many different child MVCs.  Communication between
  parent and child MVCs allows for very flexible linking of the
  MVCs.  We provide an implementation of our work in the R package
  \Rpackage{MVCClass} available at \url{http://www.bioconductor.org}. 
\end{abstract}

\section{Introduction}\label{Sec:Intro}

@@

@@

As with the model classes, the view classes need an update
method and thus, all views have an \Rfunction{updateView} method that is
defined in the \Rpackage{iSNetwork} package.  This method must take the
information that a model has just been updated and appropriately update the
view to reflect that change.  Each method depends on how the view
represents the data so for a \Robject{spreadView} object the
\Rfunction{updateView} method may select or unselect a row on the spreadsheet
while for a \Robject{sPlotView} object the \Rfunction{updateView} method may
change the appearance of a point on the plot.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{MVCMessageClass2.jpg}}
%    \scalebox{0.6}{\includegraphics{newMessageClass2.jpg}}
    \caption{ Inheritance for Message Classes that are Passed Between
      \Robject{MVCs}. }
    \label{Fig:BetwMess}
  \end{center}
\end{figure}

@@

\section*{Acknowledgements}
We are indebted to Dr. Duncan Temple Lang for the creation of the
\Rpackage{RGtk} and \Rpackage{gtkDevice} packages as well as helpful advice on
their use.
